
# SASC Workshop Package/Software Installation 

This script is intended for use to download all packages and software 
in R required for the second SASC workshop. You will need to run this 
ccript in R

## Required Software 

Before starting installation, you will need the following software version:
- R version 4.1.0 or higher
- Bioconductor version 3.14 or higher


## Installing Monocle3 

Note: This software is dependent on a lot of various packages
and may take a long installation time. 


### Install Bioconductor

Note: the Bioconductor version may be dependent on your R version. 
Change the Bioconductor version if needed. 

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install(version = "3.18")
```

**Install Bioconductor dependencies**

```{r}
BiocManager::install(c('BiocGenerics', 'DelayedArray', 'DelayedMatrixStats',
                       'limma', 'lme4', 'S4Vectors', 'SingleCellExperiment',
                       'SummarizedExperiment', 'batchelor', 'HDF5Array',
                       'terra', 'ggrastr'))
```

**Install monocle3 through cole-trapnell-lab GitHub**

```{r}
install.packages("devtools")
devtools::install_github('cole-trapnell-lab/monocle3')
```


**Test Insallation**
To ensure that Monocle3 was installed correction, run:
```{r}
library(monocle3)
```


**Monocle3 Installion troubleshooting**
If you run into errors installing Monocle3, please refer to common errors that most
users encounter on the Monocle website: 
[Monocle3 Website](https://cole-trapnell-lab.github.io/monocle3/docs/installation/)


## Installing kBET

Installion should take less than 5 min. 

There are two methods to install kBET: 

**Via Github and devtools**

```{r}
library(devtools)
install_github('theislab/kBET')
```

**Manually**

```{r}
install.packages('kBET.zip', repos = NULL, type = 'source')
```

**Test that kBET installed correction**

```{r}
library(kBET)
```


## Installing Harmony 
```{r}
install.packages("harmony")
library(harmony)
```

## Installing Misc CRAN packages
```{r}
install.packages("tidyverse")
library(tidyverse)

install.packages("randomcolorR")
library(randomcolorR)
```
