---
title: "SASC: Paired multiome, Part 2 (Kevin Lin)"
date: "June 17, 2025"
output: 
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float:
      toc_collapsed: true
    theme: united
vignette: >
  %\VignetteIndexEntry{SASC: Paired multiome, Part 2 (Kevin Lin)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
set.seed(10)
```

# Loading packages
  
```{r, eval = FALSE}
library(Seurat)
library(Signac)
library(tiltedCCA)
library(ggplot2)
library(cowplot)
library(devtools)
```

```{r, echo = FALSE}
library(Seurat)
library(Signac)
library(tiltedCCA)
library(ggplot2)
library(cowplot)
library(devtools)
```

# Loading the multiome dataset

This tutorial roughly follows [https://linnykos.github.io/tiltedCCA/articles/embryo.html](https://linnykos.github.io/tiltedCCA/articles/embryo.html).
We load a dataset downloaded from [https://www.dropbox.com/s/gh2om9eb5edgi14/10x_trevino_simplified.RData](https://www.dropbox.com/s/gh2om9eb5edgi14/10x_trevino_simplified.RData).

```{r loading}
load("/Users/kevinlin/Library/CloudStorage/Dropbox/Collaboration-and-People/sumie-katie/tutorial/sasc_2025/10x_trevino_simplified.RData")

seurat_obj
```

# RNA dimension reduction

We first perform a typical dimension reduction and UMAP on the RNA modality. (SCTransform has already been used to normalize the counts.)

```{r rna1, eval = FALSE}
set.seed(10)
Seurat::DefaultAssay(seurat_obj) <- "SCT"
seurat_obj <- Seurat::ScaleData(seurat_obj)
seurat_obj <- Seurat::RunPCA(seurat_obj, 
                             verbose = FALSE)
seurat_obj <- Seurat::RunUMAP(seurat_obj, 
                              dims = 1:50, 
                              reduction.name = "umap.rna", 
                              reduction.key = "rnaUMAP_")
```


```{r rna2, echo = FALSE}
set.seed(10)
Seurat::DefaultAssay(seurat_obj) <- "SCT"
seurat_obj <- Seurat::ScaleData(seurat_obj)
seurat_obj <- Seurat::RunPCA(seurat_obj, 
                             verbose = FALSE)
seurat_obj <- Seurat::RunUMAP(seurat_obj, 
                              dims = 1:50, 
                              reduction.name = "umap.rna", 
                              reduction.key = "rnaUMAP_")
```

```{r}
seurat_obj
```


# ATAC dimension reduction

Now we do this on the ATAC modality. Note that we typically remove the first latent dimension when constructing the UMAP, as done in [https://stuartlab.org/signac/articles/pbmc_vignette#non-linear-dimension-reduction-and-clustering](https://stuartlab.org/signac/articles/pbmc_vignette#non-linear-dimension-reduction-and-clustering).

```{r atac1, eval = FALSE}
# We exclude the first dimension as this is typically correlated with sequencing depth
set.seed(10)
seurat_obj <-  Signac::RunSVD(seurat_obj)  
seurat_obj <- Seurat::RunUMAP(seurat_obj,
                              reduction = "lsi", 
                              dims = 2:50, 
                              reduction.name = "umap.atac", 
                              reduction.key = "atacUMAP_")

seurat_obj <- tiltedCCA::rotate_seurat_embeddings(
  seurat_obj = seurat_obj,
  source_embedding = "umap.rna",
  target_embedding = "umap.atac"
)
```

```{r atac2, echo = FALSE}
# We exclude the first dimension as this is typically correlated with sequencing depth
set.seed(10)
seurat_obj <-  Signac::RunSVD(seurat_obj)  
seurat_obj <- Seurat::RunUMAP(seurat_obj,
                              reduction = "lsi", 
                              dims = 2:50, 
                              reduction.name = "umap.atac", 
                              reduction.key = "atacUMAP_")

seurat_obj <- tiltedCCA::rotate_seurat_embeddings(
  seurat_obj = seurat_obj,
  source_embedding = "umap.rna",
  target_embedding = "umap.atac"
)
```

(As a side note, to illustrate that the leading latent dimension in `lsi` is correlated with the sequencing counts, we can make the following plots.)

```{r}
lsi_mat <- seurat_obj[["lsi"]]@cell.embeddings
sequencing_depth <- seurat_obj$nCount_ATAC
cor_vec <- apply(lsi_mat, 2, function(x){stats::cor(x, sequencing_depth)})
plot(cor_vec, pch = 16, xlab = "LSI latent dimension", ylab = "Correlation w/ ATAC seq. depth")
```


```{r}
seurat_obj
```


# Plotting 

```{r color}
col_palette <- c(
  "Cyc. Prog." = rgb(213, 163, 98, maxColorValue = 255),
  "GluN2" = rgb(122, 179, 232, maxColorValue = 255),
  "GluN3"= rgb(174, 198, 235, maxColorValue = 255),
  "GluN4" = rgb(217, 227, 132, maxColorValue = 255),
  "GluN5" = rgb(127, 175, 123, maxColorValue = 255),
  "nIPC/GluN1" = rgb(114, 169, 158, maxColorValue = 255),
  "RG" = rgb(197, 125, 95, maxColorValue = 255)
)

```

Now let's visualize both modalities. This gives us a sense of what the "typical" modality looks like (i.e., no integration).

```{r plot-original1, eval = FALSE}
p1 <- Seurat::DimPlot(seurat_obj, 
                      reduction = "umap.rna",
                      group.by = "celltype", 
                      cols = col_palette, 
                      label = T, repel = T, label.size = 2.5)
p1 <- p1 + Seurat::NoLegend()
p1 <- p1 + ggplot2::ggtitle("RNA") + ggplot2::labs(x = "", y = "")
p1 <- p1 + ggplot2::theme(legend.text = ggplot2::element_text(size = 5))

p2 <- Seurat::DimPlot(seurat_obj, 
                      reduction = "umap.atac",
                      group.by = "celltype", 
                      cols = col_palette, 
                      label = T, repel = T, label.size = 2.5)
p2 <- p2 + Seurat::NoLegend()
p2 <- p2 + ggplot2::ggtitle("ATAC") + ggplot2::labs(x = "", y = "")
p2 <- p2 + ggplot2::theme(legend.text = ggplot2::element_text(size = 5))

p_all <- cowplot::plot_grid(p1, p2, ncol = 2)
p_all
```

```{r plot-original2, echo = FALSE}
p1 <- Seurat::DimPlot(seurat_obj, 
                      reduction = "umap.rna",
                      group.by = "celltype", 
                      cols = col_palette, 
                      label = T, repel = T, label.size = 2.5)
p1 <- p1 + Seurat::NoLegend()
p1 <- p1 + ggplot2::ggtitle("RNA") + ggplot2::labs(x = "", y = "")
p1 <- p1 + ggplot2::theme(legend.text = ggplot2::element_text(size = 5))

p2 <- Seurat::DimPlot(seurat_obj, 
                      reduction = "umap.atac",
                      group.by = "celltype", 
                      cols = col_palette, 
                      label = T, repel = T, label.size = 2.5)
p2 <- p2 + Seurat::NoLegend()
p2 <- p2 + ggplot2::ggtitle("ATAC") + ggplot2::labs(x = "", y = "")
p2 <- p2 + ggplot2::theme(legend.text = ggplot2::element_text(size = 5))

p_all <- cowplot::plot_grid(p1, p2, ncol = 2)
p_all
```

# WNN

We follow the tutorial in [https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis](https://satijalab.org/seurat/articles/weighted_nearest_neighbor_analysis). The intent of the Weighted Nearest Neighbor (WNN) method is to create **one** UMAP that combines the variation in either the RNA or ATAC modality (i.e., the "union" of information). This is useful because it gives us a bird's eye view of both modalities in one plot -- cell separation patterns in either RNA or ATAC would show up in the WNN UMAP.

Note that this method is not specific to RNA or ATAC modalities. It simply requies two embeddings for paired modalities (i.e., two modalities measured on the same cells).

See [https://www.cell.com/cell/fulltext/S0092-8674(21)00583-3](Hao et al., Cell, 2021) for more information.

```{r wnn1, eval = FALSE}
seurat_obj <- Seurat::FindMultiModalNeighbors(seurat_obj, 
                                              reduction.list = list("pca", "lsi"), 
                                              dims.list = list(1:50, 2:50))
seurat_obj <- Seurat::RunUMAP(seurat_obj, 
                              nn.name = "weighted.nn", 
                              reduction.name = "wnn.umap", 
                              reduction.key = "wnnUMAP_")
```

```{r wnn2, echo = FALSE}
# Identify multimodal neighbors. These will be stored in the neighbors slot, 
# and can be accessed using bm[['weighted.nn']]
# The WNN graph can be accessed at bm[["wknn"]], 
# and the SNN graph used for clustering at bm[["wsnn"]]
# Cell-specific modality weights can be accessed at bm$RNA.weight
seurat_obj <- Seurat::FindMultiModalNeighbors(seurat_obj, 
                                              reduction.list = list("pca", "lsi"), 
                                              dims.list = list(1:50, 2:50))
seurat_obj <- Seurat::RunUMAP(seurat_obj, 
                              nn.name = "weighted.nn", 
                              reduction.name = "wnn.umap", 
                              reduction.key = "wnnUMAP_")
```

```{r wnn-plot}
p1 <- Seurat::DimPlot(seurat_obj, 
              reduction = "umap.rna", 
              group.by = "celltype", 
              cols = col_palette, 
              label = TRUE, 
              label.size = 2.5, 
              repel = TRUE) + ggtitle("RNA")
p2 <- Seurat::DimPlot(seurat_obj, 
                      reduction = "umap.atac", 
                      group.by = "celltype", 
                      cols = col_palette, 
                      label = TRUE, 
                      label.size = 2.5, 
                      repel = TRUE) + ggtitle("ATAC")
p3 <- Seurat::DimPlot(seurat_obj, 
                      reduction = "wnn.umap", 
                      group.by = "celltype", 
                      cols = col_palette, 
                      label = TRUE, 
                      label.size = 2.5,
                      repel = TRUE) + ggtitle("WNN")
p1 + p2 + p3 & Seurat::NoLegend() & theme(plot.title = element_text(hjust = 0.5))
```

```{r}
seurat_obj
```

# Tilted-CCA

What if, in contrast, we want to separate which sources of variation are **shared** between the RNA and ATAC modalities, as well as which additional variations in one modalities are "invisible" to the other? This is where we could use Tilted-CCA to aid us.

We follow the tutorial from [https://linnykos.github.io/tiltedCCA/articles/embryo.html](https://linnykos.github.io/tiltedCCA/articles/embryo.html).

See [https://www.pnas.org/doi/10.1073/pnas.2303647120](Lin and Zhang, PNAS, 2024) for more information.

```{r tcca-load}
# https://linnykos.github.io/tiltedCCA/articles/embryo.html

Seurat::DefaultAssay(seurat_obj) <- "SCT"
mat_1 <- Matrix::t(seurat_obj[["SCT"]]@data)
Seurat::DefaultAssay(seurat_obj) <- "ATAC"
mat_2 <- Matrix::t(seurat_obj[["ATAC"]]@data)

mat_1b <- mat_1
sd_vec <- sparseMatrixStats::colSds(mat_1b)
if(any(sd_vec <= 1e-6)){
  mat_1b <- mat_1b[,-which(sd_vec <= 1e-6)]
}

mat_2b <- mat_2
sd_vec <- sparseMatrixStats::colSds(mat_2b)
if(any(sd_vec <= 1e-6)){
  mat_2b <- mat_2b[,-which(sd_vec <= 1e-6)]
}
```


The first step of TCCA might take a while, since this first step computes the low-dimensional embeddings in its own way.

```{r tcca-step1}
set.seed(10)
multiSVD_obj <- tiltedCCA::create_multiSVD(mat_1 = mat_1b, mat_2 = mat_2b,
                                           dims_1 = 1:50, dims_2 = 2:50,
                                           center_1 = T, center_2 = F,
                                           normalize_row = T,
                                           normalize_singular_value = T,
                                           recenter_1 = F, recenter_2 = T,
                                           rescale_1 = F, rescale_2 = T,
                                           scale_1 = T, scale_2 = F,
                                           verbose = 1)
```

The following steps are simplified to speed up the calculations for this tutorial.

```{r tcca-othersteps1, eval = FALSE}
multiSVD_obj <- tiltedCCA::form_metacells(input_obj = multiSVD_obj,
                                          large_clustering_1 = NULL, 
                                          large_clustering_2 = NULL, 
                                          num_metacells = 100,
                                          verbose = 1)

multiSVD_obj <- tiltedCCA::compute_snns(input_obj = multiSVD_obj,
                                        latent_k = 20,
                                        num_neigh = 15,
                                        bool_cosine = T,
                                        bool_intersect = F,
                                        min_deg = 15,
                                        verbose = 1)

multiSVD_obj2 <- tiltedCCA::tiltedCCA(input_obj = multiSVD_obj,
                                      discretization_gridsize = 5, # set to be lower to speed up the calculation for sake of the tutorial
                                      verbose = 1)

multiSVD_obj2 <- tiltedCCA::fine_tuning(input_obj = multiSVD_obj2,
                                        max_iter = 3, # set to be lower to speed up the calculation for sake of the tutorial
                                        verbose = 1)

multiSVD_obj2 <- tiltedCCA::tiltedCCA_decomposition(input_obj = multiSVD_obj2,
                                                    verbose = 1,
                                                    bool_modality_1_full = T,
                                                    bool_modality_2_full = F)

```

```{r tcca-othersteps2, echo = FALSE}
multiSVD_obj <- tiltedCCA::form_metacells(input_obj = multiSVD_obj,
                                          large_clustering_1 = NULL, 
                                          large_clustering_2 = NULL, 
                                          num_metacells = 100,
                                          verbose = 1)

multiSVD_obj <- tiltedCCA::compute_snns(input_obj = multiSVD_obj,
                                        latent_k = 20,
                                        num_neigh = 15,
                                        bool_cosine = T,
                                        bool_intersect = F,
                                        min_deg = 15,
                                        verbose = 1)

multiSVD_obj2 <- tiltedCCA::tiltedCCA(input_obj = multiSVD_obj,
                                      discretization_gridsize = 5, # set to be lower to speed up the calculation for sake of the tutorial
                                      verbose = 1)

multiSVD_obj2 <- tiltedCCA::fine_tuning(input_obj = multiSVD_obj2,
                                        max_iter = 3, # set to be lower to speed up the calculation for sake of the tutorial
                                        verbose = 1)

multiSVD_obj2 <- tiltedCCA::tiltedCCA_decomposition(input_obj = multiSVD_obj2,
                                                    verbose = 1,
                                                    bool_modality_1_full = T,
                                                    bool_modality_2_full = F)

```

```{r}
names(multiSVD_obj2)
```


We can now visualize the common embedding (`common_tcca`, showing the variations in both RNA and ATAC), and the two distinct embeddings (`distinct1_tcca` for the additional unique variation in RNA not in ATAC, and `distinct2_tcca` for the additional unique variation in ATAC not in RNA).

```{r tcca-umap1, eval = FALSE}
set.seed(10)
seurat_obj[["common_tcca"]] <- tiltedCCA::create_SeuratDim(input_obj = multiSVD_obj2,
                                                           what = "common",
                                                           aligned_umap_assay = "umap.rna",
                                                           seurat_obj = seurat_obj,
                                                           seurat_assay = "SCT",
                                                           verbose = 1)
set.seed(10)
seurat_obj[["distinct1_tcca"]] <- tiltedCCA::create_SeuratDim(input_obj = multiSVD_obj2,
                                                              what = "distinct_1",
                                                              aligned_umap_assay = "umap.rna",
                                                              seurat_obj = seurat_obj,
                                                              seurat_assay = "SCT",
                                                              verbose = 1)
set.seed(10)
seurat_obj[["distinct2_tcca"]] <- tiltedCCA::create_SeuratDim(input_obj = multiSVD_obj2,
                                                              what = "distinct_2",
                                                              aligned_umap_assay = "umap.rna",
                                                              seurat_obj = seurat_obj,
                                                              seurat_assay = "SCT",
                                                              verbose = 1)

```

```{r tcca-umap2, echo = FALSE}
set.seed(10)
seurat_obj[["common_tcca"]] <- tiltedCCA::create_SeuratDim(input_obj = multiSVD_obj2,
                                                           what = "common",
                                                           aligned_umap_assay = "umap.rna",
                                                           seurat_obj = seurat_obj,
                                                           seurat_assay = "SCT",
                                                           verbose = 1)
set.seed(10)
seurat_obj[["distinct1_tcca"]] <- tiltedCCA::create_SeuratDim(input_obj = multiSVD_obj2,
                                                              what = "distinct_1",
                                                              aligned_umap_assay = "umap.rna",
                                                              seurat_obj = seurat_obj,
                                                              seurat_assay = "SCT",
                                                              verbose = 1)
set.seed(10)
seurat_obj[["distinct2_tcca"]] <- tiltedCCA::create_SeuratDim(input_obj = multiSVD_obj2,
                                                              what = "distinct_2",
                                                              aligned_umap_assay = "umap.rna",
                                                              seurat_obj = seurat_obj,
                                                              seurat_assay = "SCT",
                                                              verbose = 1)

```

```{r tcca-plotting}
p1 <- Seurat::DimPlot(seurat_obj, 
                      reduction = "common_tcca",
                      group.by = "celltype", 
                      cols = col_palette, 
                      label = T, repel = T, label.size = 2.5)
p1 <- p1 + Seurat::NoLegend()
p1 <- p1 + ggplot2::ggtitle("Common") + ggplot2::labs(x = "", y = "")
p1 <- p1 + ggplot2::theme(legend.text = ggplot2::element_text(size = 5))

p2 <- Seurat::DimPlot(seurat_obj, 
                      reduction = "distinct1_tcca",
                      group.by = "celltype", 
                      cols = col_palette, 
                      label = T, repel = T, label.size = 2.5)
p2 <- p2 + Seurat::NoLegend()
p2 <- p2 + ggplot2::ggtitle("RNA Distinct") + ggplot2::labs(x = "", y = "")
p2 <- p2 + ggplot2::theme(legend.text = ggplot2::element_text(size = 5))

p3 <- Seurat::DimPlot(seurat_obj, 
                      reduction = "distinct2_tcca",
                      group.by = "celltype", 
                      cols = col_palette, 
                      label = T, repel = T, label.size = 2.5)
p3 <- p3 + Seurat::NoLegend()
p3 <- p3 + ggplot2::ggtitle("ATAC Distinct") + ggplot2::labs(x = "", y = "")
p3 <- p3 + ggplot2::theme(legend.text = ggplot2::element_text(size = 5))

p_all <- cowplot::plot_grid(p1, p2, p3, ncol = 3)
p_all
```

Equipped with this, we can visualize the **synchrony score** for each cell. This could help us understand epigenetic priming -- the pre-configuring of a cell’s chromatin so that key developmental genes are kept in a poised, low-level accessible state; this "molecular bookmark" enables rapid and lineage-specific activation (or repression) of those genes when differentiation cues arrive, thereby guiding cells toward their future identities with both speed and fidelity.

This can be computed by comparing each cell's low-dimension vector in the common embedding to the corresponding vectors in distinct embeddings. A high synchrony score means that the two modalities are "aligned" (based on some statistical definition) in both modalities, while a low score means that there is information in one modality not reflected in the other.

```{r}
synchrony_score <- tiltedCCA::compute_synchrony(input_obj = multiSVD_obj2)

seurat_obj$synchrony <- synchrony_score[Seurat::Cells(seurat_obj),"synchrony_rescaled"]

p1 <- Seurat::FeaturePlot(seurat_obj,
                    reduction = "common_tcca",
                    feature = "synchrony",
                    cols = c("forestgreen", "yellow"))
p1 <- p1 + ggplot2::ggtitle("Synchrony score") + ggplot2::labs(x = "Common TCCA 1", y = "Common TCCA 2")
p1
```

```{r}
seurat_obj
```


# Sesssion info

```{r session-info}
devtools::session_info()
```
