---
title: "SASC: Paired multiome, Part 1 (Kevin Lin)"
date: "June 17, 2025"
output: 
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float:
      toc_collapsed: true
    theme: united
vignette: >
  %\VignetteIndexEntry{SASC: Paired multiome, Part 1 (Kevin Lin)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
set.seed(10)
```

# SASC Workshop Package/Software Installation 

This script is intended for use to download all packages and software 
in R required for the second SASC workshop. You will need to run this 
ccript in R

## Required Software 

Before starting installation, it is best to have the following versions:
- R version 4.3.2 or higher
- Bioconductor version 3.18 or higher

## Install packages available on CRAN

```{r, eval = FALSE}
install.packages("ggplot2")
install.packages("cowplot")
install.packages("patchwork")
install.packages("RSQLite")
install.packages("devtools")
```

To check if the installation worked:

```{r}
library(ggplot2)
library(cowplot)
library(patchwork)
library(RSQLite)
library(devtools)
```


These next two packages could take a long time.

```{r, eval = FALSE}
install.packages("Seurat")
install.packages("Signac")
```

To check if the installation worked:

```{r}
library(Seurat)
library(Signac)
```

## Install Bioconductor

Note: the Bioconductor version may be dependent on your R version. 
Change the Bioconductor version if needed. 

```{r, eval = FALSE}
install.packages("BiocManager")
```

To check if the installation worked:

```{r}
library(BiocManager)
```

## Install packages available on BioConductor

Installing packages from BioConductor could take a long time. 
Typicall, you'll be prompted if you want to update your R packages.
If you have time to spare (typically on the order of 1 hour), you could choose to update "all".
If you're in a hurry (for example, installing during the SASC session), things typically work if you only update the CRAN packages.

```{r, eval = FALSE}
BiocManager::install(c("EnsDb.Mmusculus.v79",
                       "BSgenome.Mmusculus.UCSC.mm10",
                       "JASPAR2020",
                       "TFBSTools",
                       "ggseqlogo",
                       "biovizBase",
                       "motifmatchr"))
```

To check if the installation worked:

```{r}
library(EnsDb.Mmusculus.v79)
library(BSgenome.Mmusculus.UCSC.mm10)
library(JASPAR2020)
library(TFBSTools)
library(ggseqlogo)
library(biovizBase)
library(motifmatchr)
```

## Packages from GitHub

We'll need one last package. This package is currently available on GitHub (but not CRAN or BioConductor).

```{r, eval = FALSE}
library("devtools")
devtools::install_github("linnykos/tiltedCCA")
```

To check if the installation worked:

```{r}
library(tiltedCCA)
```

# Checking package versions

If you want to double-check the package versions which I used successfully for this tutorial:

```{r}
devtools::session_info()
```

