---
title: "SASC: Hooke Installation"
date: "April 8, 2026"
output: 
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float:
      toc_collapsed: true
    theme: united
vignette: >
  %\VignetteIndexEntry{SASC: Hooke Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(10)
```

# SASC Workshop Package/Software Installation

This script is intended to install the R packages required for the Hooke
session. You will need to run this script in R.

## Required Software

Before starting installation, it is best to have the following:

- R version 3.5.0 or higher
- A working GitHub package installation workflow in R

Hooke is currently available through GitHub. The package is also in beta, so
some features may still be under active development.

## Install packages from CRAN

These packages are needed to install the GitHub-hosted dependencies:

```{r, eval = FALSE}
install.packages("devtools")
install.packages("remotes")
install.packages("tidyverse")
install.packages("splines")
install.packages("ggh4x")
```

To check if the installation worked:

```{r, eval = FALSE}
library(devtools)
library(remotes)
library(tidyverse)
library(splines)
library(ggh4x)
```

## Install Monocle3

Hooke builds on top of `monocle3`, which also needs to be installed from
GitHub.

```{r, eval = FALSE}
devtools::install_github("cole-trapnell-lab/monocle3", ref = "develop")
```

To check if the installation worked:

```{r, eval = FALSE}
library(monocle3)
```

## Install PLNmodels

Hooke also depends on `PLNmodels`, which is installed from GitHub.

```{r, eval = FALSE}
remotes::install_github("pln-team/PLNmodels")
```

To check if the installation worked:

```{r, eval = FALSE}
library(PLNmodels)
```

## Install Hooke

Install the main `hooke` package from GitHub:

```{r, eval = FALSE}
devtools::install_github("cole-trapnell-lab/hooke")
```

To check if the installation worked:

```{r, eval = FALSE}
library(hooke)
```

## Optional: install the develop branch

If you specifically need the development branch, you can install it with:

```{r, eval = FALSE}
devtools::install_github("cole-trapnell-lab/hooke", ref = "develop")
```

## Checking package versions

If you want to double-check the package versions used in your session:

```{r, eval = FALSE}
devtools::session_info()
```
