---
title: "SASC: Hooke Part 1, Silicosis"
date: "April 8, 2026"
output:
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float:
      toc_collapsed: true
    theme: united
vignette: >
  %\VignetteIndexEntry{SASC: Hooke Part 1, Silicosis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
set.seed(10)
```


# Differential abundance

Hooke can identify differentially abundant cell types for a given contrast.
The underlying PLN model mimics the standard `R::stats` GLM-like interface.
This allows users to fit a multivariate Poisson lognormal model after
correcting for effects of offsets and covariates. This framework allows users
to perform multivariate statistical regression to describe how perturbations
alter the relative abundances of each cell state. For example, you could model
the effects of genotype while controlling for experimental batch.

## Analysis of differentially abundant cell types in silicosis with Hooke

These data were published in [Hasegawa, Franks, et al.
_bioRxiv_](https://www.biorxiv.org/content/10.1101/2023.02.17.528996v1).
The authors performed a longitudinal transcriptomic analysis using
scRNA-sequencing of lungs from pre- and post-silica challenged mice.

The data includes 12 whole lung samples across pre- and post-intratracheal
silica. Thirty-five unique cell states were identified using highly and
specifically expressed marker genes. For simplicity, samples are grouped into two levels: exposed and not exposed. The goal is to identify which cell types
are differentially abundant after silica exposure.

Download the data: [Silicosis data (.RDS)](https://depts.washington.edu/trapnell-lab/software/hooke/silicosis_cds.rds)

### Cell metadata breakdown

Important columns in this data:

* `ID` - individual sample ID. 
* `Timepoint` - timepoint at which the lung was sampled. 
* `Rep` - biological replicate
* `fine_annotation` - The finest cell type annotation level.
* `broad_annotation` - A broader level of cell type annotation than `cell_type_sub`, but still capturing all uniquely identified cell types.

The next chunk loads the silicosis dataset, creates a simplified exposure
label, and visualizes the annotated cell states.

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(monocle3)
library(hooke)
library(tidyverse)

set.seed(1)
cds <- readRDS("silicosis_cds.rds")

# For simplicity, lump together pre and post i.t. silica.
colData(cds)$exposed <- ifelse(colData(cds)$Timepoint == 0, "not exposed", "exposed")
colData(cds)$Rep <- as.factor(colData(cds)$Rep)

plot_cells(cds, color_cells_by = "fine_annotation", label_groups_by_cluster = FALSE)
```


### Making a cell count set object

The first step in Hooke is to create a `cell_count_set` object. This requires
three inputs:

- `cds`: a monocle3 `cell_data_set` object
- `sample_group`: a column in `colData(cds)` that specifies how cells are grouped into samples
- `cell_group`: a column in `colData(cds)` that specifies how cells are grouped into cell types or states

```{r, eval = TRUE, warning = FALSE, message = FALSE}
ccs <- new_cell_count_set(
    cds,
    sample_group = "ID",
    cell_group = "fine_annotation"
)
```

The resulting `ccs` object stores cell counts for each cell group within each
sample and serves as the main input for downstream Hooke modeling.

### Fitting a cell count model

The next step is to fit a `cell_count_model` given a model formula. This
function uses standard R formula notation.

A cell count model requires the following inputs:

- `ccs`: a Hooke `cell_count_set` object
- `main_model_formula_str`: a character string specifying the model of cell abundances across samples, where terms refer to columns in `colData(ccs)`
- `nuisance_model_formula_str`: a character string specifying nuisance effects across samples
- `vhat_method`: method used to estimate the variance-covariance matrix of the model coefficients.
- `num_bootstraps`: number of bootstraps to perform if `vhat_method` is "bootstrap"

In this example, the comparison of interest is silica-exposed mouse lungs
versus unexposed lungs.

```{r, eval = TRUE, warning = FALSE, message = FALSE}
ccm <- new_cell_count_model(
    ccs,
    main_model_formula_str = "~ exposed",
    vhat_method = "bootstrap",
    num_bootstraps = 10
)
```

### Estimating abundances

Next, predict abundance values for a contrast of interest using
`estimate_abundances()`.
These are model-based estimated abundances for each cell group under each
condition, not the raw observed cell counts from the dataset.

- `ccm`: a Hooke `cell_count_model` object
- `newdata`: a tibble of variables used for the prediction
- `min_log_abund`: numeric minimum log abundance value

```{r, eval = TRUE, warning = FALSE, message = FALSE}
cond_exp <- estimate_abundances(ccm, tibble::tibble(exposed = "exposed"))
cond_not_exp <- estimate_abundances(ccm, tibble::tibble(exposed = "not exposed"))

cond_not_exp %>% head()
```


### Comparing abundances

We can then compare two estimated abundance states from a Hooke model with
`compare_abundances()`. In this case, we compare exposed versus not exposed.
This step estimates differential abundance between the two predicted
conditions.

- `ccm`: a Hooke `cell_count_model` object
- `cond_x`: a cell type abundance estimate from `estimate_abundances()`
- `cond_y`: a cell type abundance estimate from `estimate_abundances()`

```{r, eval = TRUE, warning = FALSE, message = FALSE}
cond_ne_v_e_tbl <- compare_abundances(ccm, cond_not_exp, cond_exp)

cond_ne_v_e_tbl %>%
    dplyr::select(
        cell_group,
        exposed_x,
        exposed_y,
        delta_log_abund,
        delta_log_abund_se,
        delta_q_value
    ) %>%
    head()
```


### Plotting differential abundance changes on a UMAP

We can visualize these fold changes estimated in `compare_abundances()` on a UMAP using
`plot_contrast()`. Significant fold changes can be filtered with
`q_value_thresh`.
In this plot, we can identify cell groups that are enriched or depleted in exposed
lungs relative to unexposed lungs.

- `ccm`: a Hooke `cell_count_model` object
- `cond_b_vs_a_tbl`: a data frame returned by `compare_abundances()`
- `log_abundance_thresh`: numeric threshold used to select cell groups by abundance
- `q_value_thresh`: remove contrasts whose q-value exceeds this threshold

```{r, eval = TRUE, warning = FALSE, message = FALSE}
plot_contrast(ccm, cond_ne_v_e_tbl, q_value_thresh = 0.05)
```


### Controlling for batch terms

If your data contains multiple experimental batches, Hooke supports batch
correction using a nuisance model term. You must also specify the batch in the prediction data frame passed to `estimate_abundances()`. 
In this example, we control for replicate. After controlling for replicate, we see differences in the significant fold
changes.

```{r, eval = TRUE, warning = FALSE, message = FALSE}
ccm_rep <- new_cell_count_model(
    ccs,
    main_model_formula_str = "~ exposed",
    nuisance_model_formula_str = "~ Rep"
)

cond_exp_rep <- estimate_abundances(
    ccm_rep,
    newdata = tibble::tibble(exposed = "exposed", Rep = "1")
)
cond_not_exp_rep <- estimate_abundances(
    ccm_rep,
    newdata = tibble::tibble(exposed = "not exposed", Rep = "1")
)
cond_ne_v_e_tbl_rep <- compare_abundances(ccm_rep, cond_not_exp_rep, cond_exp_rep)

plot_contrast(ccm_rep, cond_ne_v_e_tbl_rep, q_value_thresh = 0.05)
```


